---
title: "Case Study Bellabeat"
author: "Akilan Sivanandham"
date: "2024-06-02"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Bellabeat, a leading high-tech manufacturer of health-focused products
for women, aims to answer key business questions by understanding and
analyzing relevant data. We will use data from the FitBit Fitness
Tracker, collected through a dispersed survey conducted from December 3,
2016, to December 5, 2016, via Amazon Mechanical Turk. Thirty Fitbit
users who met the eligibility requirements consented to submit their
personal tracker data, which includes minute-by-minute output for heart
rate, physical activity, and sleep tracking.

We will clean and manipulate this dataset to ensure accurate and
reliable analysis. Our approach will involve creating visualizations and
employing statistical models to identify trends and patterns within the
data.

The analysis will provide actionable insights for Bellabeat, offering
recommendations on how to refine its marketing strategy and enhance
product development. Ultimately, our goal is to help Bellabeat better
understand its customers and the market, enabling data-driven decisions
on the following questions.

### Our main focus on following questions

-   What are the primary patterns in Bellabeat smart device utlization?
-   How can the above insights enhance Bellabeat on its marketing
    strategy and product development?
-   What are the main opportunities and challenges that Bellabeat faces
    in the smart device industry?

### ASK

##### Task

Sršen has tasked you with analyzing smart device usage data to gain
insights into how consumers use non-Bellabeat smart devices. Based on
these insights, you are to select one Bellabeat product to focus on in
your presentation

###Prepare

#### Loading packages

In order to proceed further we need to looad following packages in our
R-Studio environment

```{r Load pakcages}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(stringr)
library(highcharter)
library(janitor)
library(RColorBrewer)
library(kableExtra)
library(scales)
library(readr)


```

#### Loading data

uploadig data in to dataset. due to number of files, we use following
method to load all the data and create dataset with appropriate names.

```{r Load data}

# Set the working directory which contain the download CSV files
setwd('Data/mturkfitbit_export_4.12.16-5.12.16/Fitabase Data 4.12.16-5.12.16')

# Get all the CSV files names from the folder.
csv_files <- list.files(pattern = '*.csv')

#Read CSV files functions
read_csv_files <- function(file_name){
  read_csv(file_name)
}

data_frame <- lapply(csv_files, read_csv_files)

#Create Data set name from the file

create_dataset_name <- function(file_name)
{
   name <- tools::file_path_sans_ext(basename(file_name))

   # Split the name by underscores
  name_parts <- unlist(strsplit(name, "_"))
  
  # Remove the last part if there are more than one part
  if (length(name_parts) > 1) {
    name_parts <- name_parts[-length(name_parts)]
  }
  
  # Combine the remaining parts
  name <- paste(name_parts, collapse = "_")
   
   
 # Insert underscores before capital letters and lowercase letters followed by capital letters
  name <- gsub("([a-z])([A-Z])", "\\1_\\2", name)    # Insert underscore between lowercase and uppercase
  name <- gsub("([A-Z])([A-Z][a-z])", "\\1_\\2", name) # Insert underscore before subsequent capital letters
  
  # Ensure all parts are capitalized correctly and join with underscores
  name_parts <- unlist(strsplit(name, "_"))
  name <- paste(name_parts, collapse = "_")
  
  name
}

# Assign data frames to dynamically created names
for (i in seq_along(csv_files)) {
  dataset_name <- create_dataset_name(csv_files[i])
  assign(dataset_name, data_frame[[i]], envir = .GlobalEnv)
}


# List the names of the created datasets
created_datasets <- sapply(csv_files, create_dataset_name)
print(created_datasets)



```

```{r check the data}
head(daily_Activity)
```

### Process

Lets check the glimpse of data, to understand the data and its type.

```{r check the Glimpse of dataset}
# Glimpse of all datasets
for (dataset_name in created_datasets) {
  cat("Dataset:", dataset_name, "\n")
  glimpse(get(dataset_name))
  cat("\n\n")
}
```

#### Split the Date and time for Datasets.

In order to process further, we need to split date and time from the
following dataset in differnt column.

-   ***heartrate_seconds*** - Dataset about the information of heart
    rate by seconds

-   ***hourly_Calories*** - Dataset about the information of Calories by
    hourly

-   ***hourly_Intensities*** - Dataset about the information of
    Intensities by hourly

-   ***hourly_Steps*** - Dataset about the information of Steps by
    hourly

-   ***sleep_Day*** - Dataset about the information of Sleeping hours by
    day

-   ***weight_Log_Info*** - Dataset about the information of Weight

```{r Modifying and adding new columns}
# heartrate_seconds
heartrate_seconds$Time = as.POSIXct(heartrate_seconds$Time,format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
heartrate_seconds$date <- format(heartrate_seconds$Time, format = "%m/%d/%y")
heartrate_seconds$time <- format(heartrate_seconds$Time, format = "%H:%M:%S")
glimpse(heartrate_seconds)
head(heartrate_seconds)

#hourly_Calories
hourly_Calories$ActivityHour = as.POSIXct(hourly_Calories$ActivityHour,format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
hourly_Calories$date <- format(hourly_Calories$ActivityHour, format = "%m/%d/%y")
hourly_Calories$time<- format(hourly_Calories$ActivityHour, format = "%H:%M:%S")
glimpse(hourly_Calories)
head(hourly_Calories)

#hourly_Intensities
hourly_Intensities$ActivityHour=as.POSIXct(hourly_Intensities$ActivityHour,format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
hourly_Intensities$date <- format(hourly_Intensities$ActivityHour, format = "%m/%d/%y")
hourly_Intensities$time<- format(hourly_Intensities$ActivityHour, format = "%H:%M:%S")
glimpse(hourly_Intensities)
head(hourly_Intensities)

#sleep_Day
sleep_Day$SleepDay=as.POSIXct(sleep_Day$SleepDay, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
sleep_Day$date <- format(sleep_Day$SleepDay, format = "%m/%d/%y")
sleep_Day$time <- format(sleep_Day$SleepDay, format = "%H:%M:%S")
glimpse(sleep_Day)
head(sleep_Day)

#hourly_Steps
hourly_Steps$ActivityHour=as.POSIXct(hourly_Steps$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
hourly_Steps$date <- format(hourly_Steps$ActivityHour, format = "%m/%d/%y")
hourly_Steps$time <- format(hourly_Steps$ActivityHour, format = "%H:%M:%S")
glimpse(hourly_Steps)
head(hourly_Steps)


#
weight_Log_Info$Date=as.POSIXct(weight_Log_Info$Date, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
weight_Log_Info$date <- format(weight_Log_Info$Date, format = "%m/%d/%y")
weight_Log_Info$time <- format(weight_Log_Info$Date, format = "%H:%M:%S")
glimpse(weight_Log_Info)
head(weight_Log_Info)

```

#### Cleaning Data

Checking for duplicates and removing it.

```{r find and remove duplicates}

# loop
for(dataset_name in created_datasets)
{
  cat("Dataset:", dataset_name, "\n")
  df <- get(dataset_name)
  
  initial_rows <- nrow(df) # get the number of rows before the process
  df <- df %>% distinct() # remove duplicates
  final_rows <- nrow(df) # get the number of rows after the process
  
  
  duplicates_removed <- initial_rows - final_rows
  
  cat("Initial rows:", initial_rows, "\n")
  
  cat("Final rows:", final_rows, "\n")
  cat("Duplicates removed:", duplicates_removed, "\n\n")

  assign(dataset_name, df, envir = .GlobalEnv) # assign the cleaned dataset to the environment
  
}

```

#### Unique data

Next we need to find number of unique data present in the dataset

```{r find the unique data}
for(dataset_name in created_datasets)
{
  cat("Dataset:", dataset_name, "\n")
  df <- get(dataset_name)
  num_unique <- n_distinct(df$Id)
  cat("Number of Unique values in 'Id':", num_unique, "\n\n")
  
}


```

From the above findings **heartrate_seconds, sleep_Day and
weight_Log_Info** have some inconsistency in data.

Especially **heartrate_seconds,** **weight_Log_Info** have only 14 and 8
unique id respectively, it can be biased data. so, this two are not
reliable to analyze further.

#### Merge data

while exploring data, Daily Activity(daily_Activity) have all the
informations except sleep data (sleep_day), so we need to combain those
data. In order to do that we need to check compactablity of column names

```{r column names}

colnames(daily_Activity)
colnames(sleep_Day)
```

From the above results, we can merge the daily_Activity with sleep_Day
using ID and Data.

```{r merge daily_Activity and sleep_Day}
# Ensure the date columns are in the same format
daily_Activity <- daily_Activity %>%
  mutate(ActivityDate = as.Date(ActivityDate, format = "%m/%d/%Y"))

# Merge the datasets on the 'Id' and corresponding date columns
daily_Activity_Sleep_merged <- merge(daily_Activity, sleep_Day, by.x = c("Id", "ActivityDate"), by.y = c("Id", "date"), all = TRUE)

# Print the merged dataset
head(daily_Activity_Sleep_merged)
```

#### Weekday calculation

```{r}
# Add weekday column and order it
weekday_Avg <- daily_Activity %>%
  mutate(weekday = weekdays(ActivityDate)) %>%
  mutate(weekday = factor(weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), ordered = TRUE))

# Calculate average daily calories and steps for each weekday
weekday_Avg <- weekday_Avg %>%
  group_by(weekday) %>%
  summarize(
    AvgDaily_Cal = mean(Calories, na.rm = TRUE),
    AvgDaily_Steps = mean(TotalSteps, na.rm = TRUE)
  )


```

### Analysis

Our data ready for analyse phase

#### Summary

Sleep summary on daily_Activity_Sleep_merged

```{r daily sleep summary}
daily_Activity_Sleep_merged %>%
  mutate(TotalHoursAsleep = TotalMinutesAsleep / 60, TotalHoursInBed = TotalTimeInBed /60) %>%
  select(TotalHoursAsleep, TotalHoursInBed) %>%
  summary()


```

From the above finding, average hours of sleep is 6.9, which is very
close to the recommended hours of sleep.

Daily Activity Summary

```{r Activity summary}


daily_Activity %>% select(LightlyActiveMinutes, FairlyActiveMinutes, VeryActiveMinutes) %>% summary()


daily_Activity %>% select( LightActiveDistance, ModeratelyActiveDistance, VeryActiveDistance) %>% summary()


daily_Activity%>% select(TotalSteps,Calories)%>% summary()

daily_Activity%>% mutate(sedentaryHours = SedentaryMinutes /60) %>%
  select(sedentaryHours, SedentaryActiveDistance)%>%summary()
```

Sedentary Hours on Average person is much higher than normal.

#### Steps taken vs Calories

```{r}

ggplot(data = daily_Activity, mapping = aes(x = TotalSteps, y = Calories)) + 
  geom_point(aes(colour = Calories), alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = FALSE, colour = "blue", linetype = "dashed") +
  scale_color_gradient(low = "yellow", high = "red") +
  labs(
    title = "Relationship Between Steps Taken and Calories Burned",
   
    x = "Total Steps",
    y = "Calories Burned",
    caption = "Source: Fitness Tracker Data"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "right"
  )


```

This visualization effectively communicates the positive relationship
between daily steps and calories burned, suggesting that increased
physical activity, measured in steps, leads to higher calorie
expenditure.

#### Compare different types of active minutes: VeryActiveMinutes, FairlyActiveMinutes, and LightlyActiveMinutes.

```{r}
active_minutes <- daily_Activity %>%
  select(Id, ActivityDate, VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes) %>%
  gather(key = "ActivityType", value = "Minutes", VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes)

ggplot(data = active_minutes, mapping = aes(x = ActivityType, y = Minutes, fill = ActivityType)) + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Comparison of Different Types of Active Minutes",
    x = "Activity Type",
    y = "Minutes",
    caption = "Source: Fitness Tracker Data"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "none"
  )
```

The bar plot compares different types of active minutes, including very
active minutes, fairly active minutes, and lightly active minutes. By
transforming the dataset, this visualization allows for a side-by-side
comparison of how different intensities of activity contribute to
overall physical activity. The plot provides insight into the
distribution of various activity levels among the users, revealing that
users spend more time in light active states compare to others.

#### Distribution of TotalDistance

```{r}
ggplot(data = daily_Activity, mapping = aes(x = TotalDistance)) + 
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Total Distance Covered",
    x = "Total Distance (miles)",
    y = "Frequency",
    caption = "Source: Fitness Tracker Data"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

This histogram displays the distribution of total distance covered by
individuals. By binning the total distance into intervals, the plot
shows the frequency of different distance ranges. This visualization
helps identify common distance ranges covered by users and highlights
the overall activity levels in terms of distance. It can reveal whether
most users tend to walk short distances daily or if there are a
significant number of users covering larger distances.

#### TotalSteps and Calories vary over ActivityDate

```{r}
daily_Activity <- daily_Activity %>%
  mutate(ActivityDate = as.Date(ActivityDate, format = "%m/%d/%Y"))

ggplot(data = daily_Activity, mapping = aes(x = ActivityDate)) + 
  geom_line(aes(y = TotalSteps, color = "Total Steps")) +
  geom_line(aes(y = Calories, color = "Calories")) +
  scale_color_manual(values = c("Total Steps" = "blue", "Calories" = "red")) +
  labs(
    title = "Daily Steps and Calories Burned Over Time",
    x = "Date",
    y = "Count",
    color = "Legend",
    caption = "Source: Fitness Tracker Data"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

The line plot shows how total steps and calories burned vary over time,
using the ActivityDate as the x-axis. By plotting both variables on the
same graph, the plot provides a temporal view of the relationship
between steps taken and calories burned. This visualization helps to
identify trends and patterns in daily activity and calorie expenditure
over time, such as seasonal variations, changes in activity levels, or
the impact of specific events. The plot underscores the consistency or
fluctuations in users' physical activity and caloric burn.

#### Weekday

Average Daily Calories Burned and Steps Taken by Weekday

```{r}
# Reshape the data for faceting
weekday_Avg_long <- weekday_Avg %>%
  pivot_longer(cols = c(AvgDaily_Cal, AvgDaily_Steps), names_to = "Metric", values_to = "Value")

# Create a faceted bar plot
ggplot(data = weekday_Avg_long, aes(x = weekday, y = Value, fill = weekday)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    title = "Average Daily Calories Burned and Steps Taken by Weekday",
    x = "Weekday",
    y = "Average Value",
    fill = "Metric",
    caption = "Source: Fitness Tracker Data"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "none",
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  )
```

The data clearly shows on Monday, Tuesday and Saturday, have more steps
taken and due to that more calories burned.

### Share

-   Daily steps and calories burned exhibit a positive correlation,
    indicating that increasing step count correlates with higher calorie
    expenditure.

-   Step count and activity intensity demonstrate a positive
    relationship, suggesting that more vigorous activity is necessary to
    achieve higher step counts.

-   There's a logical association between time spent in bed and minutes
    of sleep, implying that longer time in bed corresponds to longer
    sleep duration.

-   No discernible correlation exists between exercise intensity and
    sleep duration, indicating that heavy activity doesn't significantly
    impact sleep duration.

-   On average, peak activity levels are observed on Saturdays,
    Tuesdays, and Mondays, with individuals engaging in more vigorous
    activity early in the week and reaching maximum activity levels on
    Saturdays.

### ACT

-   BellaBeat could implement a feature to analyze users' activity
    patterns and suggest personalized strategies to increase overall
    physical activity levels, such as taking short active breaks
    throughout the day or incorporating activities into daily routines.

-   To promote better sleep quality, BellaBeat can offer personalized
    sleep hygiene tips and reminders, such as maintaining a consistent
    sleep schedule, creating a relaxing bedtime routine, and minimizing
    screen time before bed.

-   BellaBeat can provide insights into users' sedentary behavior and
    encourage them to break up prolonged sitting periods with brief
    movement breaks or light stretching exercises.

-   Incorporating amification elements, BellaBeat could introduce
    challenges and rewards to motivate users to achieve their activity
    goals, fostering a sense of accomplishment and social interaction
    within the BellaBeat community.

-   BellaBeat can leverage wearable technology to track users' heart
    rate variability (HRV) and provide personalized stress management
    recommendations, such as breathing exercises, mindfulness practices,
    or guided relaxation sessions.

-   To enhance accountability and motivation, BellaBeat could offer
    social sharing features, allowing users to connect with friends or
    join virtual fitness groups to share progress, provide
    encouragement, and participate in friendly competitions.

-   BellaBeat can integrate with third-party fitness apps and devices to
    provide a comprehensive overview of users' health and wellness data,
    enabling seamless synchronization and holistic insights into their
    lifestyle habits.

-   BellaBeat could offer educational resources and articles on topics
    related to physical activity, nutrition, sleep, stress management,
    and overall well-being to empower users with knowledge and support
    informed decision-making.
